---
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    df_print: paged
    theme: cosmo
    toc: true
    toc_float: true
---


```{r snakemake_object, include=FALSE, eval=FALSE}
if (interactive()) { ## fake data for testing
  Snakemake <- setClass("Snakemake", slots=c(input="list", output="list",
                                             params="list", threads="numeric"))
  snakemake <- Snakemake(
    input=list(h5="data/se/se_peak10k_cell10k_48FL_assays.h5",
               rds="data/se/se_peak10k_cell10k_48FL_se.rds",
               csv_trait="results/trs/scavenge_yu_2022/df_trs_covid19_knn30_bg200_rng20241106.csv.gz"),
    output=list(rmd="results/reports/trs/scavenge_yu_2022/df_trs_covid19_knn30_bg200_rng20241106.html"),
    params=list(rseed="20241106",
                trait_set_id="scavenge_yu_2022",
                trait_id="covid19"),
    threads=4L)
}

print(snakemake)
```


```{r echo=FALSE}
title_str <- sprintf("SCAVENGE %s - Full Dataset", snakemake@params$trait_id)
```

---
title: `r title_str`
---

# Purpose

Here we explore the SCAVENGE results for the trait **`r snakemake@params$trait_id`** in trait set 
**`r snakemake@params$trait_set_id`**. Analyses with initialized with seed **`r snakemake@params$rseed`**.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(SummarizedExperiment)
library(HDF5Array)
library(tidyverse)
library(magrittr)
```

## Parameters
```{r set_params}
set.seed(snakemake@params$seed)

stopifnot(dirname(snakemake@input$h5) == dirname(snakemake@input$rds))
PATH_INPUT=dirname(snakemake@input$rds)
PREFIX_INPUT=gsub("se.rds", "", basename(snakemake@input$rds))

theme_set(theme_bw())
```

# Data
## Loading
```{r load_data, message=FALSE}
se_peak_cell <- loadHDF5SummarizedExperiment(PATH_INPUT, PREFIX_INPUT)

df_trait <- read_csv(snakemake@input$csv_trait, show_col_types=FALSE)
```

## Merge Data
```{r merge_data}
mat_umap <- metadata(se_peak_cell)$reducedDims$UMAP[df_trait$cell_id,] %>% `colnames<-`(c("UMAP1", "UMAP2"))
mat_lsi <- metadata(se_peak_cell)$reducedDims$LSI[df_trait$cell_id,]

df_trait <- colData(se_peak_cell[,df_trait$cell_id]) %>%
  as_tibble(rownames="cell_id") %>%
  left_join(x=df_trait, by="cell_id") %>%
  bind_cols(mat_umap) %>% bind_cols(mat_lsi) %>%
  mutate(is_omitted=is.na(trs_raw))
```


# Cell Types
## UMAP
```{r umap_celltype, fig.width=12, fig.height=8}
df_trait %>%
  ggplot(aes(UMAP1, UMAP2, color=celltype)) +
  geom_point(size=0.1, alpha=0.1, na.rm=TRUE) +
  scale_color_discrete(guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  labs(x="UMAP 1", y="UMAP 2", color=NULL)
```

## LSI
```{r lsi_celltype, fig.width=12, fig.height=8}
df_trait %>%
  ggplot(aes(LSI1, LSI2, color=celltype)) +
  geom_point(size=0.1, alpha=0.1, na.rm=TRUE) +
  scale_color_discrete(guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  labs(x="LSI 1", y="LSI 2", color=NULL)

df_trait %>%
  ggplot(aes(LSI3, LSI4, color=celltype)) +
  geom_point(size=0.1, alpha=0.1, na.rm=TRUE) +
  scale_color_discrete(guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  labs(x="LSI 3", y="LSI 4", color=NULL)
```

# Omission Checking
#### Cell Type
```{r max_ct}
max_omit_celltype <- df_trait %>%
  group_by(celltype) %>%
  summarize(frac_omitted=mean(is_omitted)) %>%
  pull(frac_omitted) %>%
  max

ceil_omit_celltype <- max(ceiling(100*max_omit_celltype), 1)
```

No more than `r ceil_omit_celltype`% of any cell type is omitted.
```{r fig.width=6, fig.height=4}
df_trait %>%
  ggplot(aes(x=celltype, fill=is_omitted)) +
  geom_bar(position='fill') +
  scale_fill_manual(values=c("grey90", "red3")) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0)) +
  labs(x=NULL, y="Fraction of Cells", fill="Omitted")

df_trait %>%
  ggplot(aes(x=celltype, fill=is_omitted)) +
  geom_bar(position='fill') +
  scale_fill_manual(values=c("grey90", "red3")) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0)) +
  coord_cartesian(ylim=c(0, ceil_omit_celltype/100)) +
  labs(x=NULL, y="Fraction of Cells", fill="Omitted")
```

#### PCW
```{r max_pcw}
max_omit_pcw <- df_trait %>%
  group_by(PCW) %>%
  summarize(frac_omitted=mean(is_omitted)) %>%
  pull(frac_omitted) %>%
  max

ceil_omit_pcw <- max(ceiling(100*max_omit_pcw), 1)
```

No more than `r ceil_omit_pcw`% of any developmental age is omitted.
```{r fig.width=5, fig.height=4}
df_trait %>%
  ggplot(aes(x=factor(PCW), fill=is_omitted)) +
  geom_bar(position='fill') +
  scale_fill_manual(values=c("grey90", "red3")) +
  scale_y_continuous(expand=c(0,0)) +
  labs(x="PCW", y="Fraction of Cells", fill="Omitted")

df_trait %>%
  ggplot(aes(x=factor(PCW), fill=is_omitted)) +
  geom_bar(position='fill') +
  scale_fill_manual(values=c("grey90", "red3")) +
  scale_y_continuous(expand=c(0,0)) +
  coord_cartesian(ylim=c(0,ceil_omit_pcw/100)) +
  labs(x="PCW", y="Fraction of Cells", fill="Omitted")
```

#### Sample
```{r max_sample}
max_omit_sample <- df_trait %>%
  group_by(sampleID) %>%
  summarize(frac_omitted=mean(is_omitted)) %>%
  pull(frac_omitted) %>%
  max

ceil_omit_sample <- max(ceiling(100*max_omit_sample), 1)
```

No more than `r ceil_omit_sample`% of any sample is omitted.
```{r fig.width=10, fig.height=4}
df_trait %>%
  ggplot(aes(x=sampleID, fill=is_omitted)) +
  geom_bar(position='fill') +
  scale_fill_manual(values=c("grey90", "red3")) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0)) +
  labs(x="Sample", y="Fraction of Cells", fill="Omitted")

df_trait %>%
  ggplot(aes(x=sampleID, fill=is_omitted)) +
  geom_bar(position='fill') +
  scale_fill_manual(values=c("grey90", "red3")) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0)) +
  coord_cartesian(ylim=c(0,ceil_omit_sample/100)) +
  labs(x="Sample", y="Fraction of Cells", fill="Omitted")
```

#### UMAP
```{r plt_sample_umap_omitted, fig.width=8, fig.height=6}
df_trait %>%
  arrange(is_omitted) %>%
  ggplot(aes(x=UMAP1, y=UMAP2, color=is_omitted)) +
  geom_point(aes(size= is_omitted, alpha=is_omitted)) +
  scale_color_manual(values=c('grey90', 'red3'), guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  scale_alpha_manual(values=c(0.1, 1), guide='none') + scale_size_manual(values=c(0.1, 0.3), guide='none') +
  labs(x="UMAP 1", y="UMAP 2", color="Omitted")
```
#### LSI
```{r plt_sample_lsi_omitted, fig.width=8, fig.height=6}
df_trait %>%
  arrange(is_omitted) %>%
  ggplot(aes(x=LSI1, y=LSI2, color=is_omitted)) +
  geom_point(aes(size= is_omitted, alpha=is_omitted)) +
  scale_color_manual(values=c('grey90', 'red3'), guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  scale_alpha_manual(values=c(0.1, 1), guide='none') + scale_size_manual(values=c(0.1, 0.3), guide='none') +
  labs(x="LSI 1", y="LSI 2", color="Omitted")

df_trait %>%
  arrange(is_omitted) %>%
  ggplot(aes(x=LSI3, y=LSI4, color=is_omitted)) +
  geom_point(aes(size= is_omitted, alpha=is_omitted)) +
  scale_color_manual(values=c('grey90', 'red3'), guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  scale_alpha_manual(values=c(0.1, 1), guide='none') + scale_size_manual(values=c(0.1, 0.3), guide='none') +
  labs(x="LSI 3", y="LSI 4", color="Omitted")
```


# chromVAR
## Z-Scores
### UMAP
```{r umap_zscore, fig.width=10, fig.height=8}
df_trait %>%
  arrange(abs(z_score)) %>%
  ggplot(aes(UMAP1, UMAP2, color=z_score)) +
  geom_point(aes(size=abs(z_score)^3), alpha=0.9, na.rm=TRUE) +
  scale_color_viridis_c(option="B") +
  scale_size_continuous(range=c(0.01, 3), guide='none') +
  labs(x="UMAP 1", y="UMAP 2", color='Z-score')
```

### Violin
```{r plt_zscore_violin, fig.width=8, fig.height=6}
q95_top_z <- df_trait %>%
  group_by(celltype) %>%
  summarize(q95=quantile(z_score, probs=0.95)) %>%
  pull(q95) %>%
  max

df_trait %>%
  mutate(celltype_fct=reorder(celltype, -z_score, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=z_score, fill=celltype)) +
  geom_violin(draw_quantiles=c(0.25,0.5,0.75), scale='width', alpha=0.6) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  labs(x=NULL, y="Z-score")

df_trait %>%
  mutate(celltype_fct=reorder(celltype, -z_score, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=z_score, fill=celltype)) +
  geom_violin(draw_quantiles=c(0.25,0.5,0.75), scale='width', alpha=0.6) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  coord_cartesian(ylim=c(NA, q95_top_z)) +
  labs(x=NULL, y="Z-score")
```

### Boxplot
```{r plt_zscore_box, fig.width=8, fig.height=6}
df_trait %>%
  mutate(celltype_fct=reorder(celltype, -z_score, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=z_score, fill=celltype)) +
  geom_boxplot(alpha=0.6) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  labs(x=NULL, y="Z-score")

df_trait %>%
  mutate(celltype_fct=reorder(celltype, -z_score, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=z_score, fill=celltype)) +
  geom_boxplot(alpha=0.6, outliers=FALSE) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  labs(x=NULL, y="Z-score")
```

# SCAVENGE TRS
## Raw
### UMAP
```{r umap_trs_raw, fig.width=10, fig.height=8}
df_trait %>%
  arrange(trs_raw) %>%
  ggplot(aes(UMAP1, UMAP2, color=trs_raw)) +
  geom_point(aes(size=trs_raw^3), alpha=0.2, na.rm=TRUE) +
  scale_color_viridis_c(option="B") +
  scale_size_continuous(range=c(0.001, 0.1), guide='none') +
  labs(x="UMAP 1", y="UMAP 2", color='TRS (raw)')
```

### LSI
```{r lsi_trs_raw, fig.width=10, fig.height=8}
df_trait %>%
  arrange(trs_raw) %>%
  ggplot(aes(LSI1, LSI2, color=trs_raw)) +
  geom_point(aes(size=trs_raw^3), alpha=0.2, na.rm=TRUE) +
  scale_color_viridis_c(option="B") +
  scale_size_continuous(range=c(0.001, 0.1), guide='none') +
  labs(x="LSI 1", y="LSI 2", color='TRS (raw)')

df_trait %>%
  arrange(trs_raw) %>%
  ggplot(aes(LSI3, LSI4, color=trs_raw)) +
  geom_point(aes(size=trs_raw^3), alpha=0.2, na.rm=TRUE) +
  scale_color_viridis_c(option="B") +
  scale_size_continuous(range=c(0.001, 0.1), guide='none') +
  labs(x="LSI 3", y="LSI 4", color='TRS (raw)')
```

### Violin
```{r plt_trs_raw_violin, fig.width=8, fig.height=6}
df_trait %>%
  filter(!is_omitted) %>%
  mutate(celltype_fct=reorder(celltype, -trs_raw, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=trs_raw, fill=celltype)) +
  geom_violin(draw_quantiles=c(0.25,0.5,0.75), scale='width', alpha=0.6) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  labs(x=NULL, y="TRS (raw)")
```

### Boxplot
```{r plt_trs_raw_box, fig.width=8, fig.height=6}
df_trait %>%
  filter(!is_omitted) %>%
  mutate(celltype_fct=reorder(celltype, -trs_raw, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=trs_raw, fill=celltype)) +
  geom_boxplot(alpha=0.6) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  labs(x=NULL, y="TRS (raw)")
```


## Scaled
### UMAP
```{r umap_trs_scale, fig.width=10, fig.height=8}
df_trait %>%
  arrange(trs_scaled) %>%
  ggplot(aes(UMAP1, UMAP2, color=trs_scaled)) +
  geom_point(aes(size=trs_scaled^3), alpha=0.2, na.rm=TRUE) +
  scale_color_viridis_c(option="B") +
  scale_size_continuous(range=c(0.001, 0.1), guide='none') +
  labs(x="UMAP 1", y="UMAP 2", color='TRS (scaled)')
```

### LSI
```{r lsi_trs_scaled, fig.width=10, fig.height=8}
df_trait %>%
  arrange(trs_scaled) %>%
  ggplot(aes(LSI1, LSI2, color=trs_scaled)) +
  geom_point(aes(size=trs_scaled^3), alpha=0.2, na.rm=TRUE) +
  scale_color_viridis_c(option="B") +
  scale_size_continuous(range=c(0.001, 0.1), guide='none') +
  labs(x="LSI 1", y="LSI 2", color='TRS (scaled)')

df_trait %>%
  arrange(trs_scaled) %>%
  ggplot(aes(LSI3, LSI4, color=trs_scaled)) +
  geom_point(aes(size=trs_scaled^3), alpha=0.2, na.rm=TRUE) +
  scale_color_viridis_c(option="B") +
  scale_size_continuous(range=c(0.001, 0.1), guide='none') +
  labs(x="LSI 3", y="LSI 4", color='TRS (scaled)')
```

### Violin
```{r plt_trs_scaled_violin, fig.width=8, fig.height=6}
df_trait %>%
  filter(!is_omitted) %>%
  mutate(celltype_fct=reorder(celltype, -trs_scaled, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=trs_scaled, fill=celltype)) +
  geom_violin(draw_quantiles=c(0.25,0.5,0.75), scale='width', alpha=0.6) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  labs(x=NULL, y="TRS (scaled)")
```

### Boxplot
```{r plt_trs_scaled_box, fig.width=8, fig.height=6}
df_trait %>%
  filter(!is_omitted) %>%
  mutate(celltype_fct=reorder(celltype, -trs_scaled, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, y=trs_scaled, fill=celltype)) +
  geom_boxplot(alpha=0.6) +
  scale_fill_discrete(guide=guide_none()) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  labs(x=NULL, y="TRS (scaled)")
```

# Significantly Enriched Cells
### UMAP
```{r plt_umap_enriched, fig.width=8, fig.height=6}
df_trait %>%
  filter(!is_omitted) %>%
  arrange(is_enriched) %>%
  ggplot(aes(x=UMAP1, y=UMAP2, color=is_enriched)) +
  geom_point(aes(size= is_omitted, alpha=is_omitted)) +
  scale_color_manual(values=c('grey90', 'steelblue2'), guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  scale_alpha_manual(values=c(0.1, 0.3), guide='none') + scale_size_manual(values=c(0.01, 0.05), guide='none') +
  labs(x="UMAP 1", y="UMAP 2", color="Trait Enriched")
```

### LSI
```{r plt_lsi_enriched, fig.width=8, fig.height=6}
df_trait %>%
  filter(!is_omitted) %>%
  arrange(is_enriched) %>%
  ggplot(aes(x=LSI1, y=LSI2, color=is_enriched)) +
  geom_point(aes(size= is_omitted, alpha=is_omitted)) +
  scale_color_manual(values=c('grey90', 'steelblue2'), guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  scale_alpha_manual(values=c(0.1, 0.3), guide='none') + scale_size_manual(values=c(0.01, 0.05), guide='none') +
  labs(x="LSI 1", y="LSI 2", color="Trait Enriched")

df_trait %>%
  filter(!is_omitted) %>%
  arrange(is_enriched) %>%
  ggplot(aes(x=LSI3, y=LSI4, color=is_enriched)) +
  geom_point(aes(size= is_omitted, alpha=is_omitted)) +
  scale_color_manual(values=c('grey90', 'steelblue2'), guide=guide_legend(override.aes=list(size=3, alpha=1))) +
  scale_alpha_manual(values=c(0.1, 0.3), guide='none') + scale_size_manual(values=c(0.01, 0.05), guide='none') +
  labs(x="LSI 3", y="LSI 4", color="Trait Enriched")
```

### Barplot
#### Number of Cells
```{r plt_n_sig, fig.width=7, fig.height=4}
df_trait %>%
  filter(!is_omitted) %>%
  mutate(celltype_fct=reorder(celltype, -trs_scaled, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, fill=is_enriched)) +
  geom_bar(color='black', linewidth=0.3) +
  scale_fill_manual(values=c("grey80", "steelblue2")) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  scale_y_continuous(expand=c(0,0,0.1,0)) + 
  labs(x=NULL, y="Cells", fill="Trait\nEnriched")
```

#### Fraction of Cells
```{r plt_frac_sig, fig.width=7, fig.height=4}
df_trait %>%
  filter(!is_omitted) %>%
  mutate(celltype_fct=reorder(celltype, -trs_scaled, FUN=median)) %>%
  ggplot(aes(x=celltype_fct, fill=is_enriched)) +
  geom_bar(position='fill', color='black', linewidth=0.3) +
  scale_fill_manual(values=c("grey80", "steelblue2")) +
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  scale_y_continuous(expand=c(0,0)) + 
  labs(x=NULL, y="Fraction of Cells", fill="Trait\nEnriched")
```



---

# Runtime Details
## Snakemake Object
<details>
```{r print_snakemake, echo=FALSE}
snakemake
```
</details>

## Session Info
<details>
```{r sesh_info, echo=FALSE}
sessionInfo()
```
</details>

## Conda Environment
<details>
```{bash conda_info, comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  PREFIX=$(dirname $(dirname $(which R)))
  conda env export -p $PREFIX
fi
```
</details>
