---
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    df_print: paged
    theme: cosmo
    toc: true
    toc_float: true
---


```{r snakemake_object, include=FALSE, eval=FALSE}
if (interactive()) { ## fake data for testing
  Snakemake <- setClass("Snakemake", slots=c(input="list", output="list",
                                             params="list", threads="numeric"))
  snakemake <- Snakemake(
    input=list(se="results/se/subsamples/se_sample300_rng20241104.Rds",
               csv_trait_set="results/trs/df_trs_chrombpnet_kupffer_knn30_bg200_sample300_rng20241104.csv.gz"),
    output=list("results/reports/trs/report_chrombpnet_kupffer_knn30_bg200_sample300_rng20241104.html"),
    params=list(n_subsample="300",
                rseed="20241104",
                trait_set_id="chrombpnet_kupffer"),
    threads=4L)
}

print(snakemake)
```


```{r echo=FALSE}
title_str <- sprintf("SCAVENGE %s - Subsampled", snakemake@params$trait_set_id)
```

---
title: `r title_str`
---

# Purpose

Here we explore the SCAVENGE results for the trait set 
**`r snakemake@params$trait_set_id`** run on a random (seed: **`r snakemake@params$rseed`**) subset of 
`r snakemake@params$n_subsample` cells per sample.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(SummarizedExperiment)
library(tidyverse)
library(magrittr)
library(pheatmap)
```

## Parameters
```{r set_params}
set.seed(snakemake@params$seed)

theme_set(theme_bw())
```

# Data
## Loading
```{r load_data, message=FALSE}
se_peak_cell <- readRDS(snakemake@input$se)

df_trait <- read_csv(snakemake@input$csv_trait_set, show_col_types=FALSE)
```

## Merge Data
```{r merge_data}
df_umap <- metadata(se_peak_cell)$reducedDims$UMAP[unique(df_trait$cell_id),] %>% 
  `colnames<-`(c("UMAP1", "UMAP2")) %>%
  as_tibble(rownames="cell_id")
df_lsi <- metadata(se_peak_cell)$reducedDims$LSI[unique(df_trait$cell_id),] %>%
  as_tibble(rownames="cell_id")

df_trait <- colData(se_peak_cell[,unique(df_trait$cell_id)]) %>%
  as_tibble(rownames="cell_id") %>%
  left_join(x=df_trait, by="cell_id") %>%
  left_join(df_umap, by="cell_id") %>% 
  left_join(df_lsi, by="cell_id") %>%
  mutate(is_omitted=is.na(trs_raw))
```


# chromVAR
## Z-Scores
```{r summarize_zscores}
df_z <- df_trait %>%
  group_by(celltype, trait_id) %>%
  summarise(mean_z_score=mean(z_score, na.rm=TRUE),
            median_z_score=median(z_score, na.rm=TRUE),
            .groups='drop')

mat_mean_z <- df_z %>%
  dplyr::select(celltype, trait_id, mean_z_score) %>%
  pivot_wider(names_from="celltype", values_from="mean_z_score") %>%
  column_to_rownames("trait_id") %>%
  as.matrix

mat_median_z <- df_z %>%
  dplyr::select(celltype, trait_id, median_z_score) %>%
  pivot_wider(names_from="celltype", values_from="median_z_score") %>%
  column_to_rownames("trait_id") %>%
  as.matrix
```

```{r echo=FALSE}
plt_height <- 2 + nrow(mat_mean_z)/6
plt_width <- 1.5 + ncol(mat_mean_z)/6 + 0.07*max(str_width(rownames(mat_mean_z)))
```

### Mean
```{r heatmap_mean_z, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_mean_z, scale="none", cutree_rows=6, cutree_cols=8)
pheatmap(mat_mean_z, scale="row", cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_mean_z_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_mean_z), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```

### Median
```{r heatmap_median_z, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_median_z, scale="none", cutree_rows=6, cutree_cols=8)
pheatmap(mat_median_z, scale="row", cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_median_z_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_median_z), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```

# SCAVENGE TRS
## Raw
```{r summarize_trs_raw}
df_trs_raw <- df_trait %>%
  group_by(celltype, trait_id) %>%
  summarise(mean_trs_raw=mean(trs_raw, na.rm=TRUE),
            median_trs_raw=median(trs_raw, na.rm=TRUE),
            .groups='drop')

mat_mean_trs_raw <- df_trs_raw %>%
  dplyr::select(celltype, trait_id, mean_trs_raw) %>%
  pivot_wider(names_from="celltype", values_from="mean_trs_raw") %>%
  column_to_rownames("trait_id") %>%
  as.matrix

mat_median_trs_raw <- df_trs_raw %>%
  dplyr::select(celltype, trait_id, median_trs_raw) %>%
  pivot_wider(names_from="celltype", values_from="median_trs_raw") %>%
  column_to_rownames("trait_id") %>%
  as.matrix
```

### Mean
```{r heatmap_mean_trs_raw, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_mean_trs_raw, scale="none", cutree_rows=6, cutree_cols=8)
pheatmap(mat_mean_trs_raw, scale="row", cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_mean_trs_raw_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_mean_trs_raw), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```

### Median
```{r heatmap_median_trs_raw, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_median_trs_raw, scale="none", cutree_rows=6, cutree_cols=8)
pheatmap(mat_median_trs_raw, scale="row", cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_median_trs_raw_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_median_trs_raw), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```


## Scaled
```{r summarize_trs_scaled}
df_trs_scaled <- df_trait %>%
  group_by(celltype, trait_id) %>%
  summarise(mean_trs_scaled=mean(trs_scaled, na.rm=TRUE),
            median_trs_scaled=median(trs_scaled, na.rm=TRUE),
            .groups='drop')

mat_mean_trs_scaled <- df_trs_scaled %>%
  dplyr::select(celltype, trait_id, mean_trs_scaled) %>%
  pivot_wider(names_from="celltype", values_from="mean_trs_scaled") %>%
  column_to_rownames("trait_id") %>%
  as.matrix

mat_median_trs_scaled <- df_trs_scaled %>%
  dplyr::select(celltype, trait_id, median_trs_scaled) %>%
  pivot_wider(names_from="celltype", values_from="median_trs_scaled") %>%
  column_to_rownames("trait_id") %>%
  as.matrix
```

### Mean
```{r heatmap_mean_trs_scaled, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_mean_trs_scaled, scale="none", cutree_rows=6, cutree_cols=8)
pheatmap(mat_mean_trs_scaled, scale="row", cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_mean_trs_scaled_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_mean_trs_scaled), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```

### Median
```{r heatmap_median_trs_scaled, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_median_trs_scaled, scale="none", cutree_rows=6, cutree_cols=8)
pheatmap(mat_median_trs_scaled, scale="row", cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_median_trs_scaled_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_median_trs_scaled), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```


# Significantly Enriched Cells
```{r summarize_frac_enriched}
df_enriched <- df_trait %>%
  group_by(celltype, trait_id) %>%
  summarise(frac_enriched=mean(is_enriched, na.rm=TRUE),
            n_enriched=sum(is_enriched, na.rm=TRUE),
            .groups='drop')

mat_n_enriched <- df_enriched %>%
  dplyr::select(celltype, trait_id, n_enriched) %>%
  pivot_wider(names_from="celltype", values_from="n_enriched") %>%
  column_to_rownames("trait_id") %>%
  as.matrix

mat_frac_enriched <- df_enriched %>%
  dplyr::select(celltype, trait_id, frac_enriched) %>%
  pivot_wider(names_from="celltype", values_from="frac_enriched") %>%
  column_to_rownames("trait_id") %>%
  as.matrix
```

### Count
#### Raw
```{r heatmap_n_enriched_raw, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_n_enriched, cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_n_enriched_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_n_enriched), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```

#### Scaled
```{r heatmap_n_enriched_scale, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_n_enriched, scale="column", cutree_rows=6, cutree_cols=8)
```

### Fraction
#### Raw
```{r heatmap_frac_enriched_raw, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_frac_enriched, cutree_rows=6, cutree_cols=8)
```

#### Similarity
```{r heatmap_frac_enriched_sim, fig.width=plt_width, fig.height=plt_width}
pheatmap(cor(mat_frac_enriched), scale="none", cutree_rows=8, cutree_cols=8, breaks=seq(-1,1,length.out=101))
```

#### Scaled
```{r heatmap_frac_enriched_scaled_row, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_frac_enriched, scale="row", cutree_rows=6, cutree_cols=8)
```

```{r heatmap_frac_enriched_scaled_column, fig.width=plt_width, fig.height=plt_height}
pheatmap(mat_frac_enriched, scale="column", cutree_rows=6, cutree_cols=8)
```


---

# Runtime Details
## Snakemake Object
<details>
```{r print_snakemake, echo=FALSE}
snakemake
```
</details>

## Session Info
<details>
```{r sesh_info, echo=FALSE}
sessionInfo()
```
</details>

## Conda Environment
<details>
```{bash conda_info, comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  PREFIX=$(dirname $(dirname $(which R)))
  conda env export -p $PREFIX
fi
```
</details>
